Index measures for oak decline severity using phenotypic descriptors

### 2020-01-08 manuscript/manuscript.Rmd
Nathan additions to introduction

Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011.
Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both predisposing and inciting factors such as drought.
Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014].

As a results of this, there are many visual phenotypic descriptors encompassing the entirety of an oak tree that contribute to what can be defined as a declining oak tree.
A severely declining oak is most prominently characterised by very poor crown condition and numerous dead branches.
It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004].

Compounding the complexity of oak decline syndromes is the addition of different types of oak decline associated with *Quercus robur* and *Q. petraea* that are often categorised based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
Mortality in trees suffering from Chronic Oak Decline (COD) can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD.
The visual, manual assignment of decline status and severity in the field is variable and error prone due to the contribution of many factors.
Doing this accurately requires much expertise, experience, and training [@ferretti1997].

Assigning discrete decline status labels neglects the true spectrum of decline severity encountered in the field.
There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline spectrum [@pontius2014].
This has the potential to greatly improve monitoring to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices.
The recently published Action Oak Knowledge Review highlighted monitoring long term consistently as a key requirement for safeguarding UK oak health in the future [@quine2019].
Coupled with this, monitoring methods should take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.

@@

## Checks

```{r check,comment='',cache=FALSE}
manuscript_check(checks = c('word count','spelling'))
```

>>>>

Trees and forests across the globe face threats from a changing climate, notably warming and an increase in drought episodes affect the health status of individual trees causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015]. 
Within this context biotic agents (Pests and Diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
Monitoring vegetation changes and understanding the underlying processes affecting tree health are important first steps to predicting the scale and extent of these threats [@hartmann2018] and plan future policies for planting and mitigation efforts [@bradford2017]. 
Current ‘omics and remote sensing approaches provide powerful tools to document and understand these changes across multiple scales (from cellular to tree to landscape scales), but an important prerequisite is the ability to accurately describe tree health status, so that even subtle changes can be documented [@dungey2018; @hornero2020]. 
Phenotyping, documenting the physical traits of an individual or group, is thus a crucial first step relevant to studies across disciplines and further, it is necessary to understand how these measurements combine to describe health status. 
Here, we present a method using unsupervised machine learning (Random Forest) to quantify both the extent and form of decline observed in oak forests across England.

Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011; @petr2014; @brown2018].
Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both predisposing and inciting factors such as drought.
Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].

Oak decline syndromes are wide ranging in cause and attemps have been made to subdivide the types of decline associated with *Quercus robur* and *Q. petraea* based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
Mortality in trees suffering from COD can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD.
The visual, manual assignment of decline status and severity in the field is variable and error prone due to the contribution of many factors.
Doing this accurately requires much expertise, experience, and training [@ferretti1997].

As a result of this, there are many visual phenotypic descriptors that contribute to categorizing an oak tree as being in decline. 
When the decline is severe there are clear symptoms visible in the crown, such as low foliage density and numerous dead branches. 
It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004], however these symptoms can develop slowly (especially in the case of COD) and simple classes are unlikely to capture the full detail of the subtle transition from being healthy to declined.

Assigning discrete decline status labels neglects the true spectrum of decline severity encountered in the field.
There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline spectrum [@pontius2014].
This has the potential to greatly improve monitoring, enable the early detection of decline for accurate forecasting and inform forestry management practices.
The recently published Action Oak Knowledge Review highlighted monitoring long term consistently as a key requirement for safeguarding UK oak health in the future [@quine2019].
Coupled with this, monitoring methods should take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.

@@

```{r check,comment='',cache=FALSE}
# manuscript_check(checks = c('word count','spelling'))
```
